---
title: "Trajectory inference on epidermal lineage"
author: "WD"
date: "13/3/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("Seurat")
library("SeuratDisk")
library("SingleCellExperiment")
library("slingshot")
library("tidyverse")
library("clusterExperiment")
library("RColorBrewer")
library("tradeSeq")
library("epoch")
library("ggplotify")
library("patchwork")
```

```{r}
denyer <- LoadH5Seurat("../intermediate_files/denyer_wt_sctransform_annotated.h5seurat")
```

# Trajectory inference on epidermal and cortex lineage cells

```{r}
identities = c(
  "Meristematic 1",
  "Atrichoblast 1",
  "Atrichoblast 2",
  "Trichoblast 1",
  "Trichoblast 2",
  "Meristematic 2",
  "Cortex 1",
  "Cortex 2", 
  "QC"
)
epidermal_lineage <- denyer[, (denyer@meta.data$Brady.annotation %in% identities)]
epidermal_lineage[["epidermal_lineage"]] <- Idents(epidermal_lineage)
```

## Get Lineage Data

```{r}
# copy object
set.seed(1234)
# epidermal_lineage_rerun <- RunPCA(epidermal_lineage, features = VariableFeatures(epidermal_lineage))

# epidermal_lineage_rerun <- RunUMAP(epidermal_lineage, dims = 1:30)

dimred <- epidermal_lineage@reductions$umap@cell.embeddings
clustering <- epidermal_lineage$epidermal_lineage
counts <- as.matrix(epidermal_lineage@assays$SCT@counts[epidermal_lineage@assays$SCT@var.features, ])
```

## Define cell lineages

```{r}
set.seed(1234)
epidermal_sce <- as.SingleCellExperiment(epidermal_lineage)
epidermal_sce <- getLineages(reducedDims(epidermal_sce)$UMAP, 
                             clusterLabels = clustering, 
                             start.clus = "QC", 
                             end.clus = c("Trichoblast 2", "Atrichoblast 1, Cortex 1"))
```

Get principal curves and convert to SlingshotDataset

```{r}
epidermal_sce <-
  getCurves(
    epidermal_sce,
    approx_points = 300,
    thresh = 0.01,
    stretch = 0.8,
    allow.breaks = FALSE,
    shrink = 0.99
  )
epidermal_sds <- SlingshotDataSet(epidermal_sce)
```

Plot lineages on 2D UMAP embedding. 

```{r}
plot(epidermal_sce@reducedDim, col = brewer.pal(9,'Set1')[clustering], pch=16, asp = 1)
lines(epidermal_sds, lwd=2, type = 'curves', col = 'black')
```

Better in ggplot

```{r}
dfs <- lapply(seq_along(slingCurves(epidermal_sce)), function(l){
    crv <- slingCurves(epidermal_sce)[[l]]
    df <- data.frame(crv$s[crv$ord, ])
    df$curve <- l
    return(df)
})
curvedata <- do.call(rbind, dfs)
```

```{r}
p <- data.frame(epidermal_sce@reducedDim) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2)) + geom_point(aes(color = clustering)) + geom_path(data = data.frame(curvedata), aes(x = UMAP_1, y = UMAP_2, group = curve)) + theme_void()

p
```

Color according to pseudotime and split by curve

```{r}
weights <- slingCurveWeights(epidermal_sds) %>%
  data.frame() %>% rownames_to_column("cell") %>%
  pivot_longer(cols = "curve1":"curve4",
               names_to = "weight",
               values_to = "value")

pseudo <- slingPseudotime(epidermal_sds) %>%
  data.frame() %>% rownames_to_column("cell") %>%
  pivot_longer(cols = "curve1":"curve4",
               names_to = "time",
               values_to = "value")

d <- epidermal_sce@reducedDim %>%
  data.frame() %>% rownames_to_column("cell")

curvedata <- data.frame(curvedata) %>%
  mutate(curve = paste0("curve", curve)) %>%
  rename("weight" = "curve")


p <- merge(weights,
      pseudo,
      by.x = c("cell", "weight"),
      by.y = c("cell", "time")) %>%
  merge(d, by = "cell") %>%
  filter(weight != "curve3") %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point(aes(color = value.y)) +
  geom_path(data = curvedata %>% filter(weight != "curve3")) +
  facet_wrap( ~ weight, labeller = labeller(
    weight = c(
      "curve1" = "Trichoblast",
      "curve2" = "Atrichoblast",
      "curve3" = NA,
      "curve4" = "Cortex"
    )
  )) +
  theme_void() +
  scale_color_viridis(
    discrete = F,
    na.value = "lightgrey",
    option = "D",
    direction = -1
  ) +
  labs(color = "Pseudotime")

p
```

Try to plot avg lineage weight on umap

```{r}
w <- weights %>% 
  filter(weight != "curve3") %>%
  group_by(cell) %>% summarize(Entropy(value))
colnames(w) <- c("cell", "avg")

merge(w,
      d,
      by = "cell") %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point(aes(color = avg)) + theme_void() + 
  scale_color_viridis(
    discrete = F,
    na.value = "lightgrey",
    option = "B",
    direction = 1
  )
```

## Interactive 3D Plotly

```{r}
dfs <- lapply(seq_along(slingCurves(epidermal_sce)), function(l){
    crv <- slingCurves(epidermal_sce)[[l]]
    df <- data.frame(crv$s[crv$ord, ])
    df$curve <- l
    return(df)
})
curvedata <- do.call(rbind, dfs)
```

```{r}
head(curvedata)
```

Plot in 3D using plotly (rerun UMAP for 3D embedding in order to do this.)

```{r}
# plot.data <- FetchData(object = epidermal_lineage, vars = c("UMAP_1", "UMAP_2", "UMAP_3", "epidermal_lineage"))
# 
# # Make a column of row name identities (these will be your cell/barcode names)
# plot.data$label <- paste(rownames(plot.data))
# 
# fig2 <- curvedata %>% group_by(curve) %>%
#   plot_ly(data = ., 
#               x = ~UMAP_1, y=~UMAP_2, z=~UMAP_3,
#               type = "scatter3d", mode="lines", marker=list(size=3)) %>% add_trace(data = plot.data, 
#         x = ~UMAP_1, y = ~UMAP_2, z = ~UMAP_3, 
#         color = ~epidermal_lineage,
#         type = "scatter3d", 
#         mode = "markers", 
#         marker = list(size = 5, width=2, alpha=0.6), # controls size of points
#         text=~label, #This is that extra column we made earlier for which we will use for cell ID
#         hoverinfo="text")
```

Keep working with these lineages. 

## DE genes along pseudotime using tradeSeq

We have to start by choosing an appropriate number of knots.

```{r}
set.seed(1234)
epidermal_K <- evaluateK(counts = counts, sds = epidermal_sds, k = 3:10, nGenes = 100,
                   verbose = TRUE, plot = TRUE)
```

Fit gene-wise NB-GAMs using nknots = 6.

```{r}
cellweights <- slingCurveWeights(epidermal_sds)
epidermal_tradeseq <- fitGAM(counts = counts,
              sds = epidermal_sds,
              cellWeights = cellweights,
              nknots=6, verbose = TRUE)
#saveRDS(epidermal_tradeseq, file = "../intermediate_files/epidermal_tradeseq.rds")
```

```{r}
table(rowData(epidermal_tradeseq)$tradeSeq$converged)
```

```{r}
plotGeneCount(epidermal_sds, models = epidermal_tradeseq, clusters = clustering)
```

Find markers for differentiated trichoblast and atrichoblast cells

```{r}
batch_one <- startVsEndTest(epidermal_tradeseq, lineages = TRUE, pseudotimeValues = c(8, 10))
```

```{r}
b1_l1 <- batch_one[, 1:10] %>% 
  filter(pvalue_lineage1 < 0.05) %>% 
  merge(t, by = 0) %>%
  rename("annotation" = ".") %>% 
  rename("Gene" = "Row.names") %>%
  arrange(desc(waldStat_lineage1))
```


fetch gene annotation. 

```{r}
library(org.At.tair.db)
x <- org.At.tairGENENAME
mapped_tairs <- mappedkeys(x)
xx <- as.list(x[mapped_tairs])
t <- unlist(xx) %>% as.data.frame()
```

```{r}
tmp <- progenitorDE %>% 
  filter(pvalue < 0.01) %>% 
  merge(t, by = 0) %>% 
  rename("annotation" = ".") %>% 
  rename("Gene" = "Row.names") %>%
  arrange(desc(waldStat))
tmp
```

Do GO enrichment on these genes

```{r}
patGO(data.frame(cluster = rep(1, times = nrow(tmp)), 
                 row.names = tmp$Gene),
      countbias,
      groups_bp, 
      0.05)
```


Plot smoothers of auxin associated processes:

```{r}
yhatSmooth <- predictSmooth(epidermal_tradeseq, gene = pseudo.signif.genes, nPoints = 50)

yhatSmooth %>%
  filter(lineage != 3) %>%
  filter(gene %in% c("AT1G51190", "AT2G21050", "AT2G33310", "AT2G46990", "AT4G30080", "AT5G17300", "AT5G47370", "AT5G57420")) %>%
  group_by(lineage) %>%
  ggplot(aes(x = time, y = yhat, color = as.factor(lineage))) + 
  geom_path() + 
  theme_cowplot() + 
  facet_wrap(~gene, scale = 'free_y')
```

Find genes associated with bifurcation of trichoblast and atrichoblast cells

```{r}
bifurcationDE <- earlyDETest(epidermal_tradeseq, knots = c(2, 4))
bifurcationDE %>% 
  filter(pvalue < 0.01) %>% 
  merge(t, by = 0, all = F) %>% 
  rename("annotation" = ".") %>% 
  rename("Gene" = "Row.names") %>%
  arrange(desc(waldStat)) %>%
  View()
```

```{r}
pseudot_association <- associationTest(epidermal_tradeseq, lineages = TRUE)
signif_pseudot_association <- pseudot_association %>% filter(pvalue < 0.01)

trich.signif.genes <- pseudot_association %>% filter(pvalue_1 < 0.01) %>% rownames_to_column("ID") %>% pull(ID)
atrich.signif.genes <- pseudot_association %>% filter(pvalue_2 < 0.01) %>% rownames_to_column("ID") %>% pull(ID)
pseudo.signif.genes <- intersect(trich.signif.genes, atrich.signif.genes)
```